Relationships in the extremes and their influence on competition and coexistence

### 2022-03-02 paper/Paper.Rnw
fixing in text citation order

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

@@

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal.  Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$. The value of $\ddot{\rho}_i$ is then determined 
in one of 
four alternative ways: $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with the language of copula statistics (see, e.g., \citealp{Ghosh_2020_AER} and \citealp{Genest_2007}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created in an analogous way.

@@

We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2019, Ellner_2016}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. The varables in (a) are ``left-tail 
associated,'' those in (b) are symmetrically associated, and those in (c) are ``right-tail associated''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these very different 
patterns of association. We used asymmetrically tail associated cases (a, c) which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Ghosh_2020_AER,Nelsen2006}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Jasmin, I appologize for the following pedantic request, I explain below why it is necessary. Can you please change panel
%labels on this and all figure from the format "A", "B", "C" to the format "(a)", "(b)", "(c)" (i.e., lower case panel labels
%with parentheses around them). The reason for this is that this is Ecology Letters style. If we do not follow their style then their
%copy editors will make the change for us. I would prefer not to have copy editors messing around with our figures. Though Ecology Letters
%copy editors are usually pretty good, I have had many bad experiences with copy editors, generally, and I have had at least a few bad
%experiences with Ecology Letters copy editors. What happens is, they introduce errors into a paper while typesetting it, and one has to
%check their work very carefully to avoid those mistakes going to press. It's hard to blame them since it must be the most boring job 
%in the world, but anyway we do not want to have errors in our published paper as a result. The best way to prevent them introducing 
%errors into our figures is to not given them any reason to make any changes at all to our figures. The same change has to be 
%made to all main text and sup mat figures, and then all callouts to figures have to also be so changed, please. I will do these
%as much as possible while I edit, but if you could please also check it will help, and I am relying on you to change the panel labels
%in the figured themselves. I am sorry again for the tedium, here. Maybe some day they will improve the publication process so we do not 
%have to worry about nonsense like this.
%***JASMIN: changed labels of figures. Done.

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Narawani_2013, Descamps_2005}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{deValpine_2014, Anderson_2019, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

@@

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal.  Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$. The value of $\ddot{\rho}_i$ is then determined 
in one of 
four alternative ways: $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with the language of copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created in an analogous way.

@@

We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2016, Ellner_2019}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. The varables in (a) are ``left-tail 
associated,'' those in (b) are symmetrically associated, and those in (c) are ``right-tail associated''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these very different 
patterns of association. We used asymmetrically tail associated cases (a, c) which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Nelsen2006,Ghosh_2020_AER}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Jasmin, I appologize for the following pedantic request, I explain below why it is necessary. Can you please change panel
%labels on this and all figure from the format "A", "B", "C" to the format "(a)", "(b)", "(c)" (i.e., lower case panel labels
%with parentheses around them). The reason for this is that this is Ecology Letters style. If we do not follow their style then their
%copy editors will make the change for us. I would prefer not to have copy editors messing around with our figures. Though Ecology Letters
%copy editors are usually pretty good, I have had many bad experiences with copy editors, generally, and I have had at least a few bad
%experiences with Ecology Letters copy editors. What happens is, they introduce errors into a paper while typesetting it, and one has to
%check their work very carefully to avoid those mistakes going to press. It's hard to blame them since it must be the most boring job 
%in the world, but anyway we do not want to have errors in our published paper as a result. The best way to prevent them introducing 
%errors into our figures is to not given them any reason to make any changes at all to our figures. The same change has to be 
%made to all main text and sup mat figures, and then all callouts to figures have to also be so changed, please. I will do these
%as much as possible while I edit, but if you could please also check it will help, and I am relying on you to change the panel labels
%in the figured themselves. I am sorry again for the tedium, here. Maybe some day they will improve the publication process so we do not 
%have to worry about nonsense like this.
%***JASMIN: changed labels of figures. Done.
